# Working with data in R

Biodiversity Informatics (BIOL 475/575)

February 16, 2022

Programmer: Sydney

In this program, we are subsetting and indexing in R and organize raw data


### Header



```{r }
# Load Libraries
library(ezknitr)

# Clear Environment & Set Seed
remove(list=ls())
set.seed(71587)
```

_____________________________________________________________________________
## 1. Finishing up basic R tasks

### Subsetting and indexing in R

Create a vector, x

```{r }
x <- c(56, 95.3, 0.4, 2.3, 4)
```

Add the second value of x to 4.7


```{r }
x[2] + 4.7 #square bracket is an index

#use c() to create a string to add multiple values

x[c(2,3)] + 4.7

x[2:3] + 4.7 #this is a more simple way to create a string
```

What is the fourth value of x?


```{r }
x[4]
```

Add names for x


```{r }
names(x) <- c("banana", "coconut", "blueberry", "strawberry", "kiwi")
```

Use the names to add the coconut's weight to 4.7


```{r }
x["coconut"] + 4.7
```

Use the names to remove the non-tropical fruit


```{r }
# One way:

nontropical <- x[c("banana","coconut", "kiwi")]

# Another way:

nontropical2 <- x[c(-3, -4)]
```

Sort the values of x by size (ascending) and print those, but don't overwrite x



```{r }
sort(x)
```

Sort the value of x by size (descending) and overwrite x with this new order


```{r }
(x <- x[order(x, decreasing = TRUE)])
```

Print the *logical* values of x where x is larger than 1


```{r }
x > 1
```

Use the same approach but now change any values less than 1 to *NA*


```{r }
ifelse(test = x < 1, 
    yes = NA, 
    no = x)
```

Calculate the mean of the fruit that are less than 50 but more than 2


```{r }
mean(x[x < 50 & x > 2], na.rm = T)
```

### Practice exercises

1. Create a new vector named "evens" that includes all even numbers between 1 and 11.

```{r }
x <- c(2, 4, 6, 8, 10)
even <- x
```

2. Create a new vector called "odds" by adding one to the "evens" vector

```{r }
y <- (x + 1)
odd <- y
```

3. Determine if "evens" is a Numeric, Integer, Character, or Logical vector type
4. Change "evens" to a different vector type, making sure to show the results

_____________________________________________________________________________
## 2. Working with data in R

*Note:* Normally, we would load in all necessary data at the top of the script,
rather than down a ways. All libraries, necessary data, etc should be loaded 
in one place for reproducibility and for better troubleshooting.

Load in the dataset:


```{r }
df.ex <- read.csv(file = "data/raw/Dataset_S1.txt")
```

Look at the structure of the data


```{r }
str(df.ex)
```

Note that the strings get loaded as factors by default. Change this:


```{r }
df.ex <- read.csv(file = "data/raw/Dataset_S1.txt", stringsAsFactors = F)
```

View head (n = 3)



```{r }
head(df.ex, n = 3)
```

Dimensions of a data frame come in "rows, columns"



```{r }
dim(df.ex)
nrow(df.ex)
ncol(df.ex)
```

Query the column names for this dataset


```{r }
colnames(df.ex)
```

Note that some column names don't make sense, change "X.GC" to "percent.GC"


```{r }
colnames(df.ex) <- ifelse(test = colnames(df.ex) == "X.GC",
                          yes = "percent.GC",
                          no = colnames(df.ex))
```

Use $ to access a single column. Specifically, calculate the average of the depth
column


```{r }
mean(df.ex$depth)
```

Now use subsetting square brackets to do the same thing:


```{r }
mean(df.ex[ , "depth"])
```

Now, calculate the average of the depth column values, but only when depth 
is greater than 5. (Hint, we have to use subsetting again here, but to subset
only rows where depth > 5.)


```{r }
mean(df.ex[df.ex$depth > 5, "depth"])
```

While not very reproducible, let's just calculate the mean of the first 10
rows of the depth column. This time do it both with the $ operator AND with 
square brackets only.


```{r }
mean(df.ex[1:10 , "depth"])

mean(df.ex$depth[1:10])
```

Add on a new column that is a test (TRUE/FALSE) of whether the genetic window 
is in the centromere location (25,800,000 to 29,700,000).


```{r }
df.ex$centromere <- NA
cent.start <- 25800000
cent.end <- 29700000


df.ex$centromere <- df.ex$start >= cent.start & df.ex$end <= cent.end

df.ex$centromere <- df.ex$start >= cent.start & df.ex$end <= cent.end
```

Tally up the results using table()


```{r }
table(df.ex$centromere)
```

Tally up the results using sum()


```{r }
sum(df.ex$centromere)
```




### Footer

spin this with:
ezspin(file = "Sydney/programs/20220216_data_in_R.R",out_dir = "Sydney/output", fig_dir = "figures",keep_md = FALSE, keep_rmd = FALSE)
